The role of relevance, competence and priors for scalar implicatures

### 2021-12-15 writing/ELM2/abstract.tex
fix typos in overview, final abstract tweaks

In our web-based rating study, participants read background stories which were designed to vary in terms of the strength of the three factors (\texttt{high} or \texttt{low} with respect to: prior probability$\times$competence$\times$ relevance, for each trigger), manipulated within-subjects.
On critical trials, participants were asked to rate three sentences on a scale ranging from ``certainly true'' to ``certainly false'' ($\widehat{=}$0--100), one per factor. 
The story ended with one of the characters in the story making an utterance containing ``some'' or ``or''. Participants then had to indicate the probability of an SI-enriched paraphrase of that utterance. 
We thus obtained judgements on the contextual factors, and on the robustness of the SI (s. \url{https://tinyurl.com/3ru9sdja} for experiment details). 
Based on the literature, we expected higher likelihood ratings for the SI-enriched paraphrase if the alternative was perceived as highly relevant, the speaker was judged as highly competent, and the stronger alternative was viewed as a priori unlikely [2,4,6]. Each participant saw four stories per trigger, sampled from 32 stories/triggers, randomly shuffled with eight structurally similar attention checks and comprehension questions.

%(71  excluded following preregistered exclusion criteria).
We analysed data from 206 participants recruited on Prolific. Their ratings were z-scored within each factor by-participant. 
We regressed the implicature likelihood ratings against the fixed effects of trigger, all factor ratings within-story, all interactions and maximal  random effects, using a Bayesian linear mixed effects model. 
%We analysed the results using a Bayesian linear mixed effects model, regressing the implicature likelihood rating against the fixed effects of all factor ratings within-story, the effect of trigger, all interactions and maximal tractable random effects. 
Participants' factor ratings by-story agreed well with the designed classification of the stories (Fig.\ref{by-item-ratings}, red~vs.~blue color on x-axis).
%Consistent with predictions of the SI account, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true (probability of the effect of prior being smaller than 0.05: $P = 0.999$). 
As predicted, participants were more likely to derive the SIs of ``some'' and ``or'' when judging speaker competence as high ($P=0.999$ for ``some'',  $P=0.993$ for ``or'' for effect sizes being $>0.05$, Fig.\ref{posteriors} for all results). 
They were also more likely to derive the SI of ``some'' when judging the prior probability of ``all'' as low ($P= 1$ for effect $<0.05$), which was not the case for ``or''. %Interestingly, we did not observe a credible effect of prior probability in the case of ``or''.
Finally, we did not observe credible effects of relevance for either trigger.
%For the trigger ``or'', we only find a positive effect of competence ($P =  0.993$). 
Given the unexpected absence of prior effects for ``or'', we computed exploratory pairwise correlations of all predictors. While no correlations were found for ``some'', we found a significant correlation between the explanatory factors prior and relevance for ``or'' ($R^2=-0.106, p<0.01$). %, and a significant correlation between competence and relevance ($R^2 = 0.127, p=0.001$) for ``or''. 
An exploratory model comparison of two models, one containing the relevance effect over one containing the prior effect, each combined with competence, revealed mild evidence in favor of the prior as a better explanatory factor than relevance (Bayes Factor $ = 0.27$). %This might be due to the difficulty of accessing relevance experimentally.
While supporting the view that SIs for both triggers rely on epistemic reasoning affected by speaker competence, our results indicate that prior and relevance might be closely connected for ``or''.
Ultimately, our results suggest that the interdependence of the three factors is more complex than just the sum of the effects anticipated in the literature. 
Our study thus outlines further directions for investigating how exactly they enter into people’s decision as to whether or not to derive an SI.

>>>>

In our web-based rating study, participants read background stories which were designed to vary in terms of the strength of the three factors (\texttt{high} or \texttt{low} with respect to: prior probability$\times$competence$\times$ relevance, for each trigger), manipulated within-subjects.
On critical trials, participants were asked to rate three sentences on a scale ranging from ``certainly true'' to ``certainly false'' ($\widehat{=}$0--100), one per factor. 
The story ended with one of the characters in the story making an utterance containing ``some'' or ``or''. Participants then had to indicate the probability of an SI-enriched paraphrase of that utterance. 
We thus obtained judgements on the contextual factors, and on the robustness of the SI (see \url{https://tinyurl.com/3ru9sdja} for experiment details). 
Based on the literature, we expected higher likelihood ratings for the SI-enriched paraphrase if the alternative was perceived as highly relevant, the speaker was judged as highly competent, and the stronger alternative was viewed as a priori unlikely [2,4,6]. Each participant saw four stories per trigger, sampled from 32 stories/triggers, randomly shuffled with eight structurally similar attention checks and comprehension questions.

%(71  excluded following preregistered exclusion criteria).
We analysed data from 206 participants recruited on Prolific. Their ratings were z-scored within each factor by-participant. 
We regressed the implicature likelihood ratings against the fixed effects of trigger, all factor ratings within-story, all interactions and maximal  random effects, using a Bayesian linear mixed effects model. 
%We analysed the results using a Bayesian linear mixed effects model, regressing the implicature likelihood rating against the fixed effects of all factor ratings within-story, the effect of trigger, all interactions and maximal tractable random effects. 
Participants' factor ratings by-story agreed well with the designed classification of the stories (Fig.~\ref{by-item-ratings}, red~vs.~blue color on x-axis).
%Consistent with predictions of the SI account, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true (probability of the effect of prior being smaller than 0.05: $P = 0.999$). 
As predicted, participants were more likely to derive the SIs of ``some'' and ``or'' when judging speaker competence as high ($P=0.999$ for ``some'',  $P=0.993$ for ``or'' for effect sizes being $>0.05$, Fig.~\ref{posteriors} for all results). 
They were also more likely to derive the SI of ``some'' when judging the prior probability of ``all'' as low ($P= 1$ for effect $<0.05$), which was not the case for ``or''. %Interestingly, we did not observe a credible effect of prior probability in the case of ``or''.
Finally, we did not observe credible effects of relevance for either trigger.
%For the trigger ``or'', we only find a positive effect of competence ($P =  0.993$). 
Given the unexpected absence of prior effects for ``or'', we computed exploratory pairwise correlations of all predictors. While no correlations were found for ``some'', we found a significant correlation between the explanatory factors prior and relevance for ``or'' ($R^2=-0.106, p<0.01$). %, and a significant correlation between competence and relevance ($R^2 = 0.127, p=0.001$) for ``or''. 
An exploratory model comparison of two models, one containing the relevance effect over one containing the prior effect, each combined with competence, revealed mild evidence in favor of the prior as a better explanatory factor than relevance (Bayes Factor $ = 3.70$). %This might be due to the difficulty of accessing relevance experimentally.
While supporting the view that SIs for both triggers rely on epistemic reasoning affected by speaker competence, our results indicate that prior and relevance might be closely connected for ``or''.
Ultimately, our results suggest that the interdependence of the three factors is more complex than just the sum of the effects anticipated in the literature, and provide further insights into how exactly they might enter into people’s decision as to whether or not to derive an SI.
